Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-05-09 manuscript/revision2/revision2.qmd
manuscript: add response to reviews folder within manuscript/revision2/ and move intermediate files to intermediate_files folder

@@

\firstpage[15]{2}

@@

Genotypes from the Gulf and Midwest subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, Gulf vegetative growth occurred before Midwestern vegetative growth, and Gulf flowering occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both Gulf vegetative growth and flowering occurred after Midwest vegetative growth and flowering. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest vegetative growth occurred over the same period, and Gulf flowering occurred after Midwestern flowering (@fig-map A). These patterns led to strong negative phenotypic correlations for the timing of vegetative growth between the North and Texas gardens, particularly in the Gulf and across all individuals (hereafter, 'Both' subpopulations), and contributed to positive phenotypic correlations for the timing of flowering that had larger magnitudes at more northern gardens (@fig-map B).

@@

We next looked for evidence of GxWeather by jointly re-estimating genetic effects across all eight common gardens using the mash model, which can flexibly capture many modes of covariance across gardens (@fig-map D, SI Appendix, Datasets 1-6). We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1, SI Appendix, Section S2). For example, the more similar that the amount of rainfall prior to flowering is for genotypes A and B between gardens in MI and MO, the nearer these covariances will be to one. If rainfalls are uncorrelated, covariance will be near zero; if there is rank changing (e.g. high rainfall before flowering of A at MO but not MI, and high rainfall before flowering for B at MI but not MO), this covariance will be near negative one. We specified multiple covariance structures for the canonical, data-driven, and GxWeather categories, each of which represented a specific pattern of covariance in genetic effect size estimates that could be included in the model. The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, they could be partially or completely captured by the covariance structures specified in the mash model.

@@

Five GxWeather covariance structures were selected by the greedy algorithm, zero to two per subpopulation: two for the start of vegetative growth, and three for flowering date. Of these five, three had mass on them in mash models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf vegetative growth fell on a covariance matrix constructed using the covariance of daylength 14 days prior to the date of vegetative growth. The covariance matrix for this weather cue was very similar to the pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). Mash models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C).

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither covariance matrix was selected as significantly improving the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had a non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six mash models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and garden-specific effects present only at one garden.

We next characterized the pairwise patterns of effects where we were confident in the sign of the effect at both gardens. If the overall pattern of phenotypic expression differs between a pair of gardens, this difference may be comprised of SNP effects that differ in sign between these gardens (effects with antagonistic pleiotropy), SNP effects that differ in magnitude - that are large in one garden or region and smaller in others, and SNP effects that are indistinguishable in the two gardens (@fig-effects A). We used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for antagonistic pleiotropy, or a sign change between conditions, carry an equal statistical burden to those for effects with the same sign.

@@

Identifying the environmental cues that are predictive of, or even correlated with, plant phenotypic responses remains a major challenge to studies interrogating gene action across many natural environments. The GxWeather photoperiod and cumulative rainfall cues we identify here are functions of the genotypes measured and capture only a minority of SNP effects on flowering. We know still less about the overwintering parameters that cause variation in the start of vegetative growth. We could only assign SNP effects to a GxWeather covariance structures in four of the six phenotype & genetic subpopulations we modeled. More generally, it is difficult to predict the time scales over which individuals may integrate environmental cues, particularly in perennial species which may integrate these cues over longer time scales. If this integration time itself varies between individuals, we cannot select a covariance structure that reflects this, though these structures would likely be highly correlated with GxWeather structures we did include. Our approach offers an opportunity to specify multiple environmental cues and compete them to explain patterns of genetic effects, allowing us to detect how important these cues are genome-wide, and how strongly each cue influences each SNP. This is a key development to further improve our understanding of genetic variation in GxE.

@@

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. None: Effects not distinguishable by magnitude nor sign of the effect. b) The fraction of effects with each GxE type for the start of vegetative growth (Greenup Date) and reproductive growth (Flowering Date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, while Texas gardens are within the natural range of the Gulf subpopulation. ](Figure_3_GxE.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% mash tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% mash tail for flowering date for each subpopulation. a) QTL mapping for the start of vegetative growth (Greenup Date), and three weather-related functions of greenup date. b) QTL mapping for the start of reproductive growth (Flowering Date), and two weather-related functions of flowering date. ](Figure_4_QTL_Overlaps.png){#fig-qtl}

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our common gardens, we used multivariate adaptive shrinkage (*mash*) on SNP effect estimates from the diversity panel (35). To do this, we first conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) for the timing of vegetative growth and flowering, using the switchgrassGWAS package and the methods in [@lovell2021]. We used the effect estimates for single nucleotide polymorphisms (SNPs) from these GWAS as preliminary effect estimates for both traits at each garden. We then jointly modeled genetic effect estimates, and the mixture of ways these effects might covary, across all eight common gardens, using a two-part procedure. We first modeled genetic effects with different combinations of covariance structures, using the effect size estimates from a subset of [19K]{.underline} relatively unlinked (r^2^ \< 0.2), randomly selected SNPs used in the GWAS. In these models (hereafter referred to as "random"), we then specified a variety of potential covariance structures among genetic effect size estimates from each garden (as expanded on below) and used a greedy algorithm to iteratively select and add covariance structures that significantly improved the log likelihood of the *mash* model. Given the high correlation between some of the covariance patterns we tested, this approach allowed us to select only the subset of patterns of covariance that significantly improved the fit of the jointly inferred structure of genetic effect estimates present in the data. We then selected 19K relatively unlinked (r^2^ \< 0.2) SNPs from each univariate GWAS that had the largest effect estimates in each GWAS (hereafter "strong" effects), which gave sets of [**48K - 80K**]{.underline} SNPs, and jointly re-estimated these effect size estimates at all eight sites using the set of significant covariance matrices that most improved the model fit on the random effects. Notably, the majority of these "strong" effects were not significant in univariate GWAS.

>>>>

pdf-engine: pdflatex

@@

\firstpage[17]{2}

@@

Genotypes from the Gulf and Midwest subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, Gulf vegetative growth occurred before Midwestern vegetative growth, and Gulf flowering occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both Gulf vegetative growth and flowering occurred after Midwest vegetative growth and flowering. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest vegetative growth occurred over the same period, and Gulf flowering occurred after Midwestern flowering (@fig-map A). These patterns led to strong negative phenotypic correlations for the timing of the start of vegetative growth between the North and Texas gardens, particularly in the Gulf and across all individuals from the Gulf and Midwest subpopulations (hereafter, 'Both' subpopulations), and contributed to positive phenotypic correlations for the timing of flowering that had larger magnitudes at more northern gardens (@fig-map B).

@@

We next looked for evidence of GxWeather by jointly re-estimating genetic effects across all eight common gardens using the mash model, which can flexibly capture many modes of covariance across gardens (@fig-map D, SI Appendix, Datasets 1-6). We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1, SI Appendix, Section S2). For example, the more similar that the amount of rainfall prior to flowering is for genotypes A and B between gardens in MI and MO, the nearer these covariances will be to one. If rainfall amounts are uncorrelated, covariance will be near zero; if there is rank changing (e.g. high rainfall before flowering of A at MO but not MI, and high rainfall before flowering for B at MI but not MO), this covariance will be near negative one. We specified multiple covariance structures for the canonical, data-driven, and GxWeather categories, each of which represented a specific pattern of covariance in genetic effect size estimates that could be included in the model. The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, they could be partially or completely captured by the covariance structures specified in the mash model.

@@

We used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood. Five GxWeather covariance structures were selected, zero to two per subpopulation: two for the start of vegetative growth, and three for flowering date. Of these five, three had mass on them in mash models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf vegetative growth fell on a covariance matrix constructed using the covariance of daylength 14 days prior to the date of vegetative growth. The covariance matrix for this weather cue was very similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). Mash models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither of these covariance matrices significantly improved the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six mash models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); in these five models, the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and garden-specific effects present only at one garden.

We next characterized the pairwise patterns of effects where we were confident in the sign of the effect at both gardens. If the overall pattern of phenotypic expression differs between a pair of gardens, this difference may be comprised of SNP effects that differ in sign between these gardens (effects with antagonistic pleiotropy), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in the two gardens (@fig-effects A). We used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for antagonistic pleiotropy, or a sign change between conditions, carry an equal statistical burden to those for effects with the same sign.

@@

Identifying the environmental cues that are predictive of, or even correlated with, plant phenotypic responses remains a major challenge to studies interrogating gene action across many natural environments. The GxWeather photoperiod and cumulative rainfall cues we identify here are functions of the genotypes measured and capture only a minority of SNP effects on flowering. We could only assign SNP effects to a GxWeather covariance structures in four of the six phenotype & genetic subpopulations we modeled. It's likely we did not include some GxWeather important to these phenological cues - for example, overwintering parameters that might cause variation in the start of vegetative growth. More generally, it is difficult to predict the time scales over which individuals may integrate environmental cues, particularly in perennial species which may integrate these cues over longer time scales. If this integration time itself varies between individuals, the covariance structures we modeled cannot reflect this, though these structures would likely be highly correlated with GxWeather structures we did include. Our approach offers an opportunity to specify multiple environmental cues and compete them to explain patterns of genetic effects, allowing us to detect how important these cues are genome-wide, and how strongly each cue influences each SNP. This is a key development to further improve our understanding of genetic variation in GxE.

@@

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. None: Effects not distinguishable by magnitude nor sign of the effect. b) The fraction of effects with each GxE type for the start of vegetative growth (Greenup Date) and reproductive growth (Flowering Date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, while Texas gardens are within the natural range of the Gulf subpopulation.](Figure_3_GxE.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% mash tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% mash tail for flowering date for each subpopulation. a) QTL mapping for the start of vegetative growth (Greenup Date), and three weather-related functions of greenup date. b) QTL mapping for the start of reproductive growth (Flowering Date), and two weather-related functions of flowering date.](Figure_4_QTL_Overlaps.png){#fig-qtl}

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our common gardens, we used multivariate adaptive shrinkage (*mash*) on SNP effect estimates from the diversity panel (35). To do this, we first conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) for the timing of vegetative growth and flowering, using the switchgrassGWAS package and the methods in [@lovell2021]. We used the effect estimates for single nucleotide polymorphisms (SNPs) from these GWAS as preliminary effect estimates for the two traits at each garden. We then jointly modeled genetic effect estimates, and the mixture of ways these effects might covary, across all eight common gardens, using a two-part procedure. We first modeled genetic effects with different combinations of covariance structures, using the effect size estimates from a subset of [19K]{.underline} relatively unlinked (r^2^ \< 0.2), randomly selected SNPs used in the GWAS. In these models (hereafter referred to as "random"), we then specified a variety of potential covariance structures among genetic effect size estimates from each garden (as expanded on below) and used a greedy algorithm to iteratively select and add covariance structures that significantly improved the log likelihood of the *mash* model. Given the high correlation between some of the covariance patterns we tested, this approach allowed us to select only the subset of patterns of covariance that significantly improved the fit of the jointly inferred structure of genetic effect estimates present in the data. We then selected 19K relatively unlinked (r^2^ \< 0.2) SNPs from each univariate GWAS that had the largest effect estimates in each GWAS (hereafter "strong" effects), which gave sets of [**48K - 80K**]{.underline} SNPs, and jointly re-estimated these effect size estimates at all eight sites using the set of significant covariance matrices that most improved the model fit on the random effects. Notably, the majority of these "strong" effects were not significant in univariate GWAS.
